\Wxxsir methods to estimate $\rr$ in the SIR model with applications to 2009 pandemic influenza

### 2020-02-28 paper/SiM-v1/sim_v1.tex
added h1n1 data

\documentclass[12pt]{article}
\usepackage[margin=1in]{geometry}
%\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{graphicx}% http://ctan.org/pkg/graphicx
\usepackage{systeme}
%% tables
\usepackage{booktabs}
\usepackage{lscape}
\usepackage[table,xcdraw]{xcolor}
\usepackage{float}
\usepackage{url}
%% Make red text
\newcommand{\com}[1]{\textcolor{red}{ #1}}

%\linespread{2}

\usepackage[authoryear]{natbib}

\usepackage{tikz}
\tikzset{
  int/.style={circle, draw, fill=blue!20, minimum size=3em},
  init/.style={pin distance=1.2cm,pin edge={loop,thin,black}}
}
\usetikzlibrary{arrows,automata}

\pagenumbering{arabic}

\newcommand{\XX}{\ensuremath{25}} % number of methods

\newcommand{\xxsir}{\ensuremath{9} } % number of SIR methods
\newcommand{\wxxsir}{nine } % lower case number as word
\newcommand{\Wxxsir}{Nine } % capitalized word

\newcommand{\rr}{\ensuremath{\mathcal{R}_0}}

\setlength{\itemsep}{0pt}

%% tif -> png
\usepackage{epstopdf}
\epstopdfDeclareGraphicsRule{.tiff}{png}{.png}{convert #1 \OutputFile}
\AppendGraphicsExtensions{.tiff}

@@

\title{\Wxxsir methods to estimate $\rr$ in the SIR model with applications to 2009 pandemic influenza}
\author{ Shannon Gallagher$^{\dag}$, Andersen Chang$^{\ddag}$, and William F. Eddy$^{\dag}$ \\$\dag$ Department of Statistics and Data Science, Carnegie Mellon University\\ $\ddag$ Department of Statistics, Rice University}
\date{\today}
\maketitle

\begin{abstract}
For nearly a century, the initial reproduction number ($\rr$) has been used to summarize infectious diseases and to compare them to one another.  Difficulties in estimating $\rr$ arise from on both how a disease transmits through a population as well as from differences in statistical estimation methods.  We describe \wxxsir methods used to estimate $\rr$ and provide a thorough simulation study of how these methods differ in the presence of disease parameters.  As real world motivation, we analyze the 2009 outbreak of the H1N1 pandemic influenza in the USA and compare the results from our nine methods to a previous study.  We discuss the most important aspects from our results which effect the estimation of $\rr$ and provide guidelines for estimating point estimates and confidence/credible intervals to create reliable, comparable estimates of $\rr$.
  \end{abstract}

@@

Both researchers want to estimate $\rr$ and so fit models to their data sets.  The details of how the models are fit are not so important at the moment (but will be in the later sections), but to fairly compare one model to another assume both researchers use a least squares estimates to estimate the disease parameters and then use the ratio estimator (RE) as an estimate of $\rr$.  That is researcher 1 estimates $\rr$ with
\begin{align*}
  (\hat{\beta}_{XEYZ}, \hat{\gamma}_{XEYZ}, \hat{\mu}_{XEYZ} )&=\arg \min_{\beta, \gamma, \mu} \sum_{t} \left [ \left (X_{XEYZ}(t) - X(t)\right )^2  \right . \\
                                                                 &+\left ( E_{XEYZ}(t) - E(t) \right )^2  \\
  &\left . + \left ( Z_{XEYZ}(t) - Z(t) \right )^2\right ]
\end{align*}
Then the RE for $\rr^{XYZ}$ is given by Equation \ref{eq:sirls},
\begin{align*}
  \hat{\rr}^{XEYZ}= \frac{\hat{\beta}_{XEYZ}}{\hat{\gamma}_{XEYZ}}.
\end{align*}
Researcher 2 estimates $\rr$ with 
\begin{align*}
(\hat{\beta}_{XYZ}, \hat{\gamma}_{XYZ} )&=\arg \min_{\beta, \gamma} \sum_{t} \left [ \left (X_{XYZ}(t) - X(t)\right )^2 + \left ( Z_{XYZ}(t) - Z(t) \right )^2 \right ].
\end{align*}
Then the RE for $\rr^{XYZ}$ is given by Equation \ref{eq:sirls},
\begin{align*}
  \hat{\rr}^{XYZ}= \frac{\hat{\beta}_{XYZ}}{\hat{\gamma}_{XYZ}}.
\end{align*}

@@

\begin{align*}
(\hat{\beta}, \hat{\gamma} )&=\arg \min_{\beta, \gamma} \sum_{t} \left [ \left (\hat{X}(t) - X(t)\right )^2 + \left ( \hat{Z}(t) - Z(t) \right )^2 \right ]
\end{align*}
Then the ratio estimator (RE) estimate for $\rr$ is given by Equation \ref{eq:sirls},
\begin{align}\label{eq:sirls}
  \hat{\rr}= \frac{\hat{\beta}}{\hat{\gamma}}.
\end{align}

@@

%%%%%%%%%%%%%%%%%%%%%
\subsection{Log-Linear (LL)}
\cite{harko2014exact} were able to reduce the SIR model to one ODE.  From this, we can derive the following,
\begin{align}
  X(t) &=  X(0) e^{\frac{-\beta}{\gamma}\frac{Z(t)}{N}} \nonumber\\
  \log \frac{X(t)}{X(0)} &=  \frac{-\beta }{\gamma}\frac{Z(t)}{N} \nonumber\\
  \log \frac{X(t)}{X(0)} &=  -\rr \frac{Z(t)}{N}. \label{eq:harko_lin}
\end{align}

We can regress the left hand side in Eq. \ref{eq:harko_lin} on $Z(t)$ (using least squares to estimate the coefficients), the number of recovered individuals, with $\rr$ as the coefficient of $Z(t)/N$ to obtain an estimate of $\rr$,
\begin{align*}
  \hat{\rr} = -\frac{\sum_{t=0}^T \log \frac{ X(t)}{X(0)}}{\sum_{t=0}^T\frac{Z(t)}{N}}.
\end{align*}
This method says that for a one percent increase in number of recovered individuals, we expect the ratio of the number of previous susceptibles to number of new susceptibles to increase by $e^{\rr}$,
\begin{align*}
  \log \left ( X(t+1)/ X(0) \right ) - \log \left ( X(t)/X(0) \right ) &= - .01\rr\\
  \log \left ( X(t+1) \right ) - \log \left ( X(t) \right )  &=- .01\rr\\
  \log \left ( X(t) / X(t+1) \right ) &= .01\rr\\
  \frac{X(t)}{X(t+1)}  &\approx e^{\rr}.
\end{align*}

@@

In this model, the number of susceptibles at the next step, $\hat{X}(t+1)$, has a Binomial distribution based on the contacts with the current number of infectious, $\hat{Y}(t)$ and the current number of susceptibles.  That is $\hat{X}(t+1) \sim \text{Binomial}\left(\hat{X}(t), \alpha^{\hat{Y}(t)}\right)$, where $\alpha$ is the probability of avoiding infection from an infective.  One can write out the likelihood for $\alpha$ in this model, namely,
\begin{align*}
\mathcal{L}\left ( \alpha ; \textnormal{Data}\right ) \propto \prod_{t=0}^{T-1}\left ( \alpha^{\hat{Y}(t)} \right )^{\hat{X}(t+1)} \left (1- \alpha^{\hat{Y}(t)} \right )^{\hat{X}(t) - \hat{X}(t+1)}.
\end{align*}
Maximizing the likelihood yields an estimate for $\alpha$,
\begin{align*}
\hat{\alpha} = \arg \max_{\alpha} \mathcal{L}\left(\alpha; \textnormal{data} \right ).
  \end{align*}
 \cite{barbour2004} report that the estimated reproduction number is then,
\begin{align}\label{eq:r0-mc}
\hat{\rr} &= \log \left ( \frac{1}{1-\hat{\alpha}}\right ).
\end{align}

@@

Denote $\hat{X}(t), \hat{Y}(t)$, and $\hat{Z}(t)$ to be the observed number in each compartment and $X(t), Y(t)$, and $Z(t)$ to be the true, underlying model at time $t$.   The stochastic model is given by
\begin{align*}
  \hat{X}(t+1) &= \hat{X}(t) - x_t\\
  \hat{Y}(t+1) & = N - \hat{X}(t+1) - \hat{Z}(t+1), \nonumber \\
  \hat{Z}(t+1) &= \hat{Z}(t) + z_t \nonumber
\end{align*}
with $\hat{X}(0)$, $\hat{Y}(0)$, and $\hat{Z}(0)$ known. The terms $x_t$ and $z_t$ in Equation \eqref{eq:sir-noise} incorporate random behavior into the model.  Here we choose them to have the following distributions,
\begin{align*}
  x_{t} &\sim \text{Binomial}\left (\hat{X}(t), \frac{\beta Y(t)}{N} \right ) \\
  z_{t} &\sim \text{Binomial}\left (\hat{Y}(t), \gamma \right ). \nonumber
\end{align*}
Then the likelihood is, when we assuming that the next time step is dependent only on the current time step,
\begin{align*}
  \mathcal{L}\left (\beta, \gamma ; \textnormal{Data} \right ) &=
                                                  \prod_{t=0}^{T-1} \left [\left ( \begin{array}{c}
                                                            \hat{X}(t) \\
                                                            x_t
                                                            \end{array}
  \right ) \left (\beta - \frac{\beta\left (X(t) +Z(t) \right )}{N} \right )^{x_t} \left (1 - \beta +  \frac{\beta\left (X(t) + Z(t) \right )}{N} \right )^{\hat{X}(t) - x_t} \cdot \right .\\
  & \left . \left ( \begin{array}{c}
              N -\left ( \hat{X}(t) + \hat{Z}(t)\right ) \\
              z_t
            \end{array}
  \right ) \left (\gamma \right )^{z_t} \left ( 1 - \gamma\right )^{N -\left ( \hat{X}(t) + \hat{Z}(t)\right ) - z_t} \right ].
\end{align*}
The estimate for $\rr$ is then
\begin{align}\label{eq:lbe-r0}
  (\hat{\beta}, \hat{\gamma}) &= \arg \max_{\beta, \gamma} \mathcal{L}\left ( \beta, \gamma; \textnormal{Data} \right )\nonumber\\
  \rr &= \frac{\hat{\beta}}{\hat{\gamma}}
  \end{align}
  The two main differences between this model and MC are  that 1) LBE looks at the joint likelihood of the XYZ model, whereas MC only explicitly models the number of susceptibles and 2) the difference in the structure of the transition probability.  We use the delta method in conjunction with $\hat{\beta}$ and $\hat{\gamma}$ estimations from the Nelder-Mead optimization of the likelihood to estimate the stantard error of $\hat{\rr}$.
  The advantages of using this model are that our modelling assumptions are explicit and that the modelling process seeems to be reasonable.  Disadvantages include maximizing the liklihood may be more time consuming than other methods due to the need to approximate $X(t)$ and $Z(t)$ for many different values of $\beta$ and $\gamma$ and that if we are wrong about the generating process of the XYZ curves, then this model may be a very poor fit.

@@

Described by \cite{bettencourt2008} and summarized in \cite{obadia2012r0}, the sequential Bayes method is a Bayesian approach to an approximation of the classic SYR model.  The approximated SIR model assumes that the incidence at $t+1$, $J(t+1)$ has a Poisson distribution, with $\gamma$ as the  average inverse of the infectious period. In order to estimate $\rr$, we must have some idea about $\gamma$,
\begin{align*}
J(t+1) | \rr  \sim \textnormal{Poisson}( J(t) \exp \left \{  \gamma (\rr-1)\right \})
\end{align*}
Then, the posterior distribution of $\rr$ given the previous days' incidence is
\begin{align*}
  P(\rr | J_0, \dots, J_{t+1}) = \frac{P(J_{t+1} | \rr, J_0, \dots, J_t)P(\rr| J_0, \dots, J_t)}{P(J_0, \dots, J_{t+1})}.
\end{align*}
This method is sequential in that the prior distribution for $\rr$ comes from the previous day.  The initial prior for $\rr$ is assumed to be flat.  This method results in a posterior distribution from which credible intervals may be obtained.  This method assumes, initial growth in incidence to be exponential, and homogeneous mixing of populations as with any compartment model.  The advantages of this method are that of the ability to obtain an entire posterior distribution, whereas many other methods are difficult to even find an estimate of the variance.  Disadvantages include strict assumptions about the distributions and computational time required to estimate the relevant parameters.

@@

We can also use the Delta method to derive variance estimates of LMA, LMAT, LL, (which in this case is equivalent to the standard error derived from the typical linear regression framework), and LBE.  We can also use the Delta method to derive an estimate for MC, where $\hat{\sigma}$ is the estimate of the variance of the $\hat{\alpha}$,
\begin{align*}
h(\alpha) &= \log \left ( \frac{1}{1-\alpha}\right ) \\
  h^\prime(\alpha) &= \alpha - 1 \\
                     V[ \hat{\rr} ] & \approx \frac{1}{T}  \left (\frac{\hat{\sigma}}{(\hat{\alpha} - 1)^2} \right ).
\end{align*}

@@

\pagebreak

\bibliographystyle{apa}%Choose a bibliograhpic style
\bibliography{Master}

\appendix

@@

%%%%%%%%%%%%%%%%%%%%%%
\textbf{Title:}

\textbf{Author:}

\textbf{Citation:}

\textbf{Major themes:}

\textbf{Notes:}
\\
%%%%%%%%%%%%%%%%%%%%%%

\begin{figure}[h]
\begin{center}
\includegraphics[width=4in]{images/mvhw7_3c.pdf}
\end{center}
\caption{
Biplots of the different continuous variables and their correlations.}\label{fig3c}
\end{figure}

%% two pictures whoa

\begin{figure}[h]
\centering

\begin{figure}[h]
\centering

\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=1\linewidth]{images/mt_eda_cont_hists.pdf}
  \caption{Histograms of Arrival Delay and continuous covariates.  Arrival delay seems to have a right skewed distribution.  This may indicate that we will be transforming this variable later on.  After transforming Air Time and Distance by a log transformation, we don't really seem to have many outliers in our covariates.  We seem to have outliers in the CRS Dep. Time and Arrival Time; however, time is cyclical and so these are not, in fact outliers.}
  \label{hists}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=1\linewidth]{images/mt_eda_cont_hists.pdf}
  \caption{\textcolor{red}{Placeholder}}
  \label{tabs1}
\end{subfigure}
\caption{}
\end{figure}

\begin{table}
\begin{tabular}{l | a | b | a | b}
\hline
\rowcolor{LightCyan}
\mc{1}{}  & \mc{1}{x} & \mc{1}{y} & \mc{1}{w} & \mc{1}{z} \    \hline
variable 1 & a & b & c & d \    variable 2 & a & b & c & d \\ \hline
\end{tabular}
\end{table}

%%
\begin{figure}
\centering
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=1\linewidth]{images/resids_full.pdf}
  \caption{}
  \label{residsf}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=1\linewidth]{images/diags_full.pdf}
  \caption{ }
  \label{diagsf}
\end{subfigure}
\caption{}
\end{figure}

>>>>

\documentclass[AMA,demo,STIX1COL]{WileyNJD-v2}
\usepackage{moreverb}

%% big red comments
\newcommand{\com}[1]{\textcolor{red}{ #1}}

%% nice tables and sideways table
\usepackage{booktabs}
\usepackage{lscape}

%% Our custom macros
\newcommand{\XX}{\ensuremath{25}} % number of methods
\newcommand{\xxsir}{\ensuremath{9} } % number of SIR methods
s\newcommand{\wxxsir}{nine } % lower case number as word
\newcommand{\Wxxsir}{Nine } % capitalized word
\newcommand{\rr}{\ensuremath{\mathcal{R}_0}} %% r0 macro

%% diff eqs
\usepackage{systeme}

%% tikz graph
\usepackage{tikz}
\tikzset{
  int/.style={circle, draw, fill=blue!20, minimum size=3em},
  init/.style={pin distance=1.2cm,pin edge={loop,thin,black}}
}
\usetikzlibrary{arrows,automata}

\newcommand\BibTeX{{\rmfamily B\kern-.05em \textsc{i\kern-.025em b}\kern-.08em
T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

\articletype{Statistics in Medicine}%

\received{<day> <Month>, <year>}
\revised{<day> <Month>, <year>}
\accepted{<day> <Month>, <year>}

% \raggedbottom

\usepackage{tikz}
\usetikzlibrary{positioning}

\tikzset{
  int/.style={circle, draw, fill=blue!20, minimum size=3em},
  init/.style={pin distance=1.2cm,pin edge={loop,thin,black}}
}
\usetikzlibrary{arrows}

@@

\title{\Wxxsir methods to estimate $\rr$ in the SIR model with applications to 2009 pandemic influenza}

\author[1]{Shannon Gallagher*}
\author[2]{Andersen Chang}
\author[1]{William F. Eddy}

\authormark{Gallagher, Chang, and Eddy}

\address[1]{\orgdiv{Department of Statistics}, \orgname{Carnegie Mellon University}, \orgaddress{\state{PA}, \country{USA}}}
\address[1]{\orgdiv{Department of Statistics}, \orgname{Rice University}, \orgaddress{\state{TX}, \country{USA}}}

\corres{*Shannon Gallagher, Department of Statistics; Carnegie Mellon University; Pittsburgh, PA 15213. \email{sgallagh@stat.cmu.edu}}

\presentaddress{Department of Statistics; Carnegie Mellon University}

\abstract[Abstract]{For nearly a century, the initial reproduction number ($\rr$) has been used to summarize infectious diseases and to compare them to one another.  Difficulties in estimating $\rr$ arise from on both how a disease transmits through a population as well as from differences in statistical estimation methods.  We describe \wxxsir methods used to estimate $\rr$ and provide a thorough simulation study of how these methods differ in the presence of disease parameters.  As real world motivation, we analyze the 2009 outbreak of the H1N1 pandemic influenza in the USA and compare the results from our nine methods to a previous study.  We discuss the most important aspects from our results which effect the estimation of $\rr$ and provide guidelines for estimating point estimates and confidence/credible intervals to create reliable, comparable estimates of $\rr$}

\keywords{Compartment models; Agent-based models; Stochastic infectious disease modelling.}

\jnlcitation{\cname{%
\author{Gallagher, S.} and
\author{Eddy, W. F}.
\ctitle{Stochastic Equivalence of Compartment and Agent-based Models}.  September 2017.}
}

\maketitle

@@

  Both researchers want to estimate $\rr$ and so fit models to their data sets.  The details of how the models are fit are not so important at the moment (but will be in the later sections), but to fairly compare one model to another assume both researchers use a least squares estimates to estimate the disease parameters and then use the ratio estimator (RE) as an estimate of $\rr$.  That is researcher 1 estimates $\rr$ with
\begin{align*}
  (\hat{\beta}_{XEYZ}, \hat{\gamma}_{XEYZ}, \hat{\mu}_{XEYZ} )&=\textnormal{arg} \textnormal{min}_{\beta, \gamma, \mu} \sum_{t} \left [ \left (X_{XEYZ}(t) - X(t)\right )^2  \right . \\
                                                                 & + \left ( E_{XEYZ}(t) - E(t) \right )^2  \\
  & \left . + \left ( Z_{XEYZ}(t) - Z(t) \right )^2\right ]
\end{align*}
Then the RE for $\rr^{XYZ}$ is given by Equation \ref{eq:sirls},
\begin{align*}
  \hat{\rr}^{XEYZ}= \frac{\hat{\beta}_{XEYZ}}{\hat{\gamma}_{XEYZ}}.
\end{align*}
Researcher 2 estimates $\rr$ with 
\begin{align*}
(\hat{\beta}_{XYZ}, \hat{\gamma}_{XYZ} )&=\textnormal{arg} \textnormal{min}_{\beta, \gamma} \sum_{t} \left [ \left (X_{XYZ}(t) - X(t)\right )^2 + \left ( Z_{XYZ}(t) - Z(t) \right )^2 \right ].
\end{align*}
Then the RE for $\rr^{XYZ}$ is given by Equation \ref{eq:sirls},
\begin{align*}
  \hat{\rr}^{XYZ}= \frac{\hat{\beta}_{XYZ}}{\hat{\gamma}_{XYZ}}.
\end{align*}

@@

\begin{align*}
(\hat{\beta}, \hat{\gamma} )&=\textnormal{arg} \textnormal{min}_{\beta, \gamma} \sum_{t} \left [ \left (\hat{X}(t) - X(t)\right )^2 + \left ( \hat{Z}(t) - Z(t) \right )^2 \right ]
\end{align*}
Then the ratio estimator (RE) estimate for $\rr$ is given by Equation \ref{eq:sirls},
\begin{align}\label{eq:sirls}
  \hat{\rr}= \frac{\hat{\beta}}{\hat{\gamma}}.
\end{align}

@@

%%%%%%%%%%%%%%%%%%%%%
\subsection{Log-Linear (LL)}
\cite{harko2014exact} were able to reduce the SIR model to one ODE.  From this, we can derive the following, 
\begin{align}
  X(t) &=  X(0) e^{\frac{-\beta}{\gamma}\frac{Z(t)}{N}} \nonumber\\
  \textnormal{log} \frac{X(t)}{X(0)} &=  \frac{-\beta }{\gamma}\frac{Z(t)}{N} \nonumber\\
  \textnormal{log} \frac{X(t)}{X(0)} &=  -\rr \frac{Z(t)}{N}. \label{eq:harko_lin}
\end{align}

We can regress the left hand side in Eq. \ref{eq:harko_lin} on $Z(t)$ (using least squares to estimate the coefficients), the number of recovered individuals, with $\rr$ as the coefficient of $Z(t)/N$ to obtain an estimate of $\rr$,
\begin{align*}
  \hat{\rr} = -\frac{\sum_{t=0}^T \textnormal{log} \frac{ X(t)}{X(0)}}{\sum_{t=0}^T\frac{Z(t)}{N}}.
\end{align*}
This method says that for a one percent increase in number of recovered individuals, we expect the ratio of the number of previous susceptibles to number of new susceptibles to increase by $e^{\rr}$,
\begin{align*}
  \textnormal{log} \left ( X(t+1)/ X(0) \right ) - \textnormal{log} \left ( X(t)/X(0) \right ) &= - .01\rr\\
  \textnormal{log} \left ( X(t+1) \right ) - \textnormal{log} \left ( X(t) \right )  &=- .01\rr\\
  \textnormal{log} \left ( X(t) / X(t+1) \right ) &= .01\rr\\
  \frac{X(t)}{X(t+1)}  &\approx e^{\rr}.
\end{align*}

@@

In this model, the number of susceptibles at the next step, $\hat{X}(t+1)$, has a Binomial distribution based on the contacts with the current number of infectious, $\hat{Y}(t)$ and the current number of susceptibles.  That is $\hat{X}(t+1) \sim \text{Binomial}\left(\hat{X}(t), \alpha^{\hat{Y}(t)}\right)$, where $\alpha$ is the probability of avoiding infection from an infective.  One can write out the likelihood for $\alpha$ in this model, namely,
\begin{align*}
\mathcal{L}\left ( \alpha ; \textnormal{Data}\right ) \propto \prod_{t=0}^{T-1}\left ( \alpha^{\hat{Y}(t)} \right )^{\hat{X}(t+1)} \left (1- \alpha^{\hat{Y}(t)} \right )^{\hat{X}(t) - \hat{X}(t+1)}.
\end{align*}
Maximizing the likelihood yields an estimate for $\alpha$,
\begin{align*}
\hat{\alpha} = \textnormal{arg} \textnormal{max}_{\alpha} \mathcal{L}\left(\alpha; \textnormal{data} \right ).
  \end{align*}
 \cite{barbour2004} report that the estimated reproduction number is then,
\begin{align}\label{eq:r0-mc}
\hat{\rr} &= \textnormal{log} \left ( \frac{1}{1-\hat{\alpha}}\right ).
\end{align}

@@

Denote $\hat{X}(t), \hat{Y}(t)$, and $\hat{Z}(t)$ to be the observed number in each compartment and $X(t), Y(t)$, and $Z(t)$ to be the true, underlying model at time $t$.   The stochastic model is given by
\begin{align*}
  \hat{X}(t+1) &= \hat{X}(t) - x_t\\
  \hat{Y}(t+1) & = N - \hat{X}(t+1) - \hat{Z}(t+1), \nonumber \\
  \hat{Z}(t+1) &= \hat{Z}(t) + z_t \nonumber
\end{align*}
with $\hat{X}(0)$, $\hat{Y}(0)$, and $\hat{Z}(0)$ known. The terms $x_t$ and $z_t$ in Equation \eqref{eq:sir-noise} incorporate random behavior into the model.  Here we choose them to have the following distributions,
\begin{align*}
  x_{t} &\sim \text{Binomial}\left (\hat{X}(t), \frac{\beta Y(t)}{N} \right ) \\
  z_{t} &\sim \text{Binomial}\left (\hat{Y}(t), \gamma \right ). \nonumber
\end{align*}
Then the likelihood is, when we assuming that the next time step is dependent only on the current time step,
\begin{align*}
  \mathcal{L}\left (\beta, \gamma ; \textnormal{Data} \right ) &=
                                                  \prod_{t=0}^{T-1} \left [\left ( \begin{array}{c}
                                                            \hat{X}(t) \\
                                                            x_t
                                                            \end{array}
  \right ) \left (\beta - \frac{\beta\left (X(t) +Z(t) \right )}{N} \right )^{x_t} \left (1 - \beta +  \frac{\beta\left (X(t) + Z(t) \right )}{N} \right )^{\hat{X}(t) - x_t} \cdot \right .\\
  & \left . \left ( \begin{array}{c}
              N -\left ( \hat{X}(t) + \hat{Z}(t)\right ) \\
              z_t
            \end{array}
  \right ) \left (\gamma \right )^{z_t} \left ( 1 - \gamma\right )^{N -\left ( \hat{X}(t) + \hat{Z}(t)\right ) - z_t} \right ].
\end{align*}
The estimate for $\rr$ is then
\begin{align}\label{eq:lbe-r0}
  (\hat{\beta}, \hat{\gamma}) &= \textnormal{arg} \textnormal{max}_{\beta, \gamma} \mathcal{L}\left ( \beta, \gamma; \textnormal{Data} \right )\nonumber\\
  \rr &= \frac{\hat{\beta}}{\hat{\gamma}}
  \end{align}
  The two main differences between this model and MC are  that 1) LBE looks at the joint likelihood of the XYZ model, whereas MC only explicitly models the number of susceptibles and 2) the difference in the structure of the transition probability.  We use the delta method in conjunction with $\hat{\beta}$ and $\hat{\gamma}$ estimations from the Nelder-Mead optimization of the likelihood to estimate the stantard error of $\hat{\rr}$.
  The advantages of using this model are that our modelling assumptions are explicit and that the modelling process seeems to be reasonable.  Disadvantages include maximizing the liklihood may be more time consuming than other methods due to the need to approximate $X(t)$ and $Z(t)$ for many different values of $\beta$ and $\gamma$ and that if we are wrong about the generating process of the XYZ curves, then this model may be a very poor fit.

@@

Described by \cite{bettencourt2008} and summarized in \cite{obadia2012r0}, the sequential Bayes method is a Bayesian approach to an approximation of the classic SYR model.  The approximated SIR model assumes that the incidence at $t+1$, $J(t+1)$ has a Poisson distribution, with $\gamma$ as the  average inverse of the infectious period. In order to estimate $\rr$, we must have some idea about $\gamma$,
\begin{align*}
J(t+1) | \rr  \sim \textnormal{Poisson}( J(t) \textnormal{exp} \left \{  \gamma (\rr-1)\right \})
\end{align*}
Then, the posterior distribution of $\rr$ given the previous days' incidence is
\begin{align*}
  P(\rr | J_0, \dots, J_{t+1}) = \frac{P(J_{t+1} | \rr, J_0, \dots, J_t)P(\rr| J_0, \dots, J_t)}{P(J_0, \dots, J_{t+1})}.
\end{align*}
This method is sequential in that the prior distribution for $\rr$ comes from the previous day.  The initial prior for $\rr$ is assumed to be flat.  This method results in a posterior distribution from which credible intervals may be obtained.  This method assumes, initial growth in incidence to be exponential, and homogeneous mixing of populations as with any compartment model.  The advantages of this method are that of the ability to obtain an entire posterior distribution, whereas many other methods are difficult to even find an estimate of the variance.  Disadvantages include strict assumptions about the distributions and computational time required to estimate the relevant parameters.

@@

We can also use the Delta method to derive variance estimates of LMA, LMAT, LL, (which in this case is equivalent to the standard error derived from the typical linear regression framework), and LBE.  We can also use the Delta method to derive an estimate for MC, where $\hat{\sigma}$ is the estimate of the variance of the $\hat{\alpha}$,
\begin{align*}
h(\alpha) &= \textnormal{log} \left ( \frac{1}{1-\alpha}\right ) \\
  h^\prime(\alpha) &= \alpha - 1 \\
                     V[ \hat{\rr} ] & \approx \frac{1}{T}  \left (\frac{\hat{\sigma}}{(\hat{\alpha} - 1)^2} \right ).
\end{align*}

@@

@@

%  To get the journal style of heading for an appendix, mimic the following.
